---
title: "README"
output: 
  rmarkdown::github_document:
    number_sections: false
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library(knitr)
library(ClickHouseHTTP)
library(dplyr)
cranRef <- function(x){
  sprintf(
    "[%s](https://CRAN.R-project.org/package=%s): %s",
    x, x, packageDescription(x)$Title
  )
}
```

[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/ClickHouseHTTP)](https://cran.r-project.org/package=ClickHouseHTTP)
[![](https://cranlogs.r-pkg.org/badges/ClickHouseHTTP)](https://cran.r-project.org/package=ClickHouseHTTP)

*ClickHouse* (<https://clickhouse.com/>)
is an open-source, high performance columnar
OLAP (online analytical processing of queries) database management system
for real-time analytics using SQL. This *DBI* backend
relies on the 'ClickHouse' HTTP interface and support HTTPS protocol.

This package has been developed as an alternative to the
excellent [RClickhouse](https://github.com/IMSMWU/RClickhouse) to provide
secured connection through SSL using HTTPS
(unfortunately SSL connection is not yet supported by RClickhouse).

The ClickHouseHTTP R package is licensed under
[GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html).

# Installation

## From CRAN

```{r, eval=FALSE}
install.packages("ClickHouseHTTP")
```

## Dependencies

The following R packages available on CRAN are required:

```{r, echo=FALSE, results='asis'}
deps <- desc::desc_get_deps()
sdeps <- filter(deps, type %in% c("Depends", "Imports") & package!="R")
for(p in sdeps$package){
   cat(paste("-", cranRef(p)), sep="\n")
}
```

And those are suggested:

```{r, echo=FALSE, results='asis'}
wdeps <- filter(deps, type=="Suggests" & package!="R")
for(p in wdeps$package){
   cat(paste("-", cranRef(p)), sep="\n")
}
```

## From github

```{r, eval=FALSE}
devtools::install_github("patzaw/ClickHouseHTTP")
```

# Documentation

## Usage

### Connection

```{r, eval=FALSE}
library(DBI)
## HTTP connection
con <- dbConnect(
   ClickHouseHTTP::ClickHouseHTTP(), host="localhost",
   port=8123
)
## HTTPS connection (without ssl peer verification)
con <- dbConnect(
   ClickHouseHTTP::ClickHouseHTTP(), host="localhost",
   port=8443, https=TRUE, ssl_verifypeer=FALSE
)
```

### Write a table in the database

```{r, eval=FALSE}
library(dplyr)
data("mtcars")
mtcars <- as_tibble(mtcars, rownames="car")
dbWriteTable(con, "mtcars", mtcars)
```

### Query the database

```{r, eval=FALSE}
carsFromDB <- dbReadTable(con, "mtcars")
dbGetQuery(con, "SELECT car, mpg, cyl, hp FROM mtcars WHERE hp>=110")
```

By default, ClickHouseHTTP relies on the
[Apache `Arrow`](https://arrow.apache.org/) format provided by ClickHouse.
However, as described in the
[documentation](https://clickhouse.com/docs/en/interfaces/formats/#data-format-arrow),
the following types are not supported in the current implementation of
this format: *TIME32*, *FIXED_SIZE_BINARY*, *JSON*, *UUID*, *ENUM*.
The `format` argument of the `dbGetQuery()` function can be used to
rely on the *TabSeparatedWithNamesAndTypes* format.

```{r, eval=FALSE}
selCars <- dbGetQuery(
   con, "SELECT car, mpg, cyl, hp FROM mtcars WHERE hp>=110",
   format="TabSeparatedWithNamesAndTypes"
)
## Identifying the original ClickHouse data types
attr(selCars, "type")
```

### Using alternative databases stored in ClickHouse

```{r, eval=FALSE}
dbSendQuery(con, "CREATE DATABASE swiss")
dbSendQuery(con, "USE swiss")
```

The chosen database is used until the session expires.
It can also be chosen when connecting using the `dbname` argument of
the `dbConnect()` function.

The example below shows that spaces in column names are supported.
It also shows the support of R `list` using the *Array* ClickHouse type.

```{r, eval=FALSE}
data("swiss")
swiss <- as_tibble(swiss, rownames="province")
swiss <- mutate(swiss, "pr letters"=strsplit(province, ""))
dbWriteTable(
   con, "swiss", swiss,
   engine="MergeTree() ORDER BY (Fertility, province)"
)
swissFromDB <- dbReadTable(con, "swiss")
```

A table from another database can also be accessed as following:

```{r, eval=FALSE}
dbReadTable(con, SQL("default.mtcars"))
```

## Setting up a ClickHouse database using docker

### Configuration

```{sh, eval=FALSE}
CH_HOME=~/Documents/Projects/Test_CH
mkdir -p $CH_HOME
mkdir -p ${CH_HOME}/data
mkdir -p ${CH_HOME}/conf
mkdir -p ${CH_HOME}/log
```

Configuration files are shared in the `supp/ClickHouse-Conf-Files` folder in
this repository.

```{sh, eval=FALSE}
cp supp/ClickHouse-Conf-Files/users.xml ${CH_HOME}/conf/
cp supp/ClickHouse-Conf-Files/config.xml ${CH_HOME}/conf/
```

### SSL certificate

```{sh, eval=FALSE}
openssl req -subj "/CN=localhost" -new -newkey rsa:2048 -days 365 -nodes -x509 -keyout ${CH_HOME}/conf/server.key -out ${CH_HOME}/conf/server.crt
openssl dhparam -out ${CH_HOME}/conf/dhparam.pem 4096
```

### Container

The following ports are supported in the shared `config.xml` file:

- 9000: Native TCP interface (not used by ClickHouseHTTP)
- 9440: Native TCP interface wrapped in TLS (not used by ClickHouseHTTP)
- 8123: HTTP interface
- 8443: HTTPS interface

```{sh, eval=FALSE}
chmod -R a+rwx ${CH_HOME}
docker run -d --name Test_CH \
	--ulimit nofile=262144:262144 \
	--volume ${CH_HOME}/data:/var/lib/clickhouse \
	--publish=9000:9000 \
	--publish=9440:9440 \
	--publish=8123:8123 \
	--publish=8443:8443 \
	--volume ${CH_HOME}/conf/users.xml:/etc/clickhouse-server/users.xml \
	--volume ${CH_HOME}/conf/config.xml:/etc/clickhouse-server/config.xml \
	--volume ${CH_HOME}/conf/server.crt:/etc/clickhouse-server/server.crt \
	--volume ${CH_HOME}/conf/server.key:/etc/clickhouse-server/server.key \
	--volume ${CH_HOME}/conf/dhparam.pem:/etc/clickhouse-server/dhparam.pem \
	--volume ${CH_HOME}/log:/var/log/clickhouse-server \
	clickhouse/clickhouse-server:22.2.3.5
```


# Alternatives

- [RClickhouse](https://github.com/IMSMWU/RClickhouse) is another DBI backend
for the ClickHouse database. It provides basic dplyr support
by auto-generating SQL-commands using dbplyr and is based on this
[C++ Clickhouse Client](https://github.com/artpaul/clickhouse-cpp).

- [clickhouse-r](https://github.com/hannes/clickhouse-r) is another DBI
backend for the ClickHouse database relying on HTTP protocol. It provides
SSL support but without peer verification for the moment.

# Acknowledgments

This work was entirely supported by [UCB Pharma](https://www.ucb.com/) (Early
Solutions department).
